# Song Popularity Prediction using spotify dataset
## By Durga, Radika, Rohit
setwd("D:/College/FallSemester/Advanced Analytics/Project/Spotify-Song-Popularity-Prediction-master/data")
```{r}

library(tidyverse)
library(dplyr)
library(caret)


```
### Loading Spotify Data
```{r}
spotify = read.csv("D:/College/FallSemester/Advanced Analytics/Project/Spotify-Song-Popularity-Prediction-master/data/SpotifyFeatures.csv")
head(spotify)
str(spotify)
names(spotify)
```


###ensuring children's music difference is resolved
```{r}
spotify[spotify == "Children's Music"] <-'Children’s Music'
```

#Checking for missing values
```{r}
sapply(spotify,function(x) sum(is.na(x)))

```
# finding duplicate records
```{r}
sum(duplicated(spotify$track_id))

```

```{r}
duplicate=spotify[duplicated(spotify$track_id), ]
dim(duplicate)
head(duplicate)
```

```{r}
spotify[spotify["track_id"]=="6iOvnACn4ChlAw4lWUU4dd"]
spotify[spotify["track_id"]=="2XGLdVl7lGeq8ksM6Al7jT"]
```
#We see that most of the attributes of the duplicated songs are the same except for 'popularity' and 'genre'.The 'popularity' column can be aggregated since it is a numerical column but dummy variables should be created for the categorical column of 'genre'.

#generating dummy values for 'genre'

```{r}
genre_list = as.list(unique(spotify$genre))
for (g in genre_list){
  spotify[g]=as.numeric((spotify$genre==g))
}

```

```{r}
head(spotify)

```
#Aggregation of popularity score
#grouping by track_id number to get rid of duplicates and keeping the maximum values in each column.

```{r}
spotify1 = aggregate(.~ track_id,data=spotify,FUN=max)

```

```{r}
head(spotify1)

```
#Removing redundant genre column


```{r}
spotify2=subset(spotify1,select=-c(genre))
dim(spotify2)
```
#We successfully addressed the duplicates of each track by aggregating them to a single row.
#We now have 176,774 unique tracks in our dataset (down from 232,725).

# TOP 50 2019

```{r}
setwd("D:/College/FallSemester/Advanced Analytics/Project/Spotify-Song-Popularity-Prediction-master/data")
top50 <- read.csv("top50.csv", stringsAsFactors = T)


```

```{r}
summary(top50$Popularity)

```

```{r}
hist(top50$Popularity)
```

# TOP 100 2019

```{r}
setwd("D:/College/FallSemester/Advanced Analytics/Project/Spotify-Song-Popularity-Prediction-master/data")

top100 <- read.csv("spotify_top_100_2019.csv", stringsAsFactors = T)

```

# summary and histogram plot of popularity

```{r}
summary(top100$popularity)

```

```{r}
hist(top100$popularity)

```
#the scores within the range 0-25 seem like outliers. We can remove outliers from this dataset with the IQR method to get a better perspective on the data.

```{r}
#find Q1, Q3, and interquartile range for values in column A
Q1 <- quantile(top100$popularity, .25)
Q3 <- quantile(top100$popularity, .75)
IQR<-IQR(top100$popularity)

#only keep rows in dataframe that have values within 1.5*IQR of Q1 and Q3
Top100_no_outliers <- subset(top100, top100$popularity> (Q1 - 1.5*IQR) & top100$popularity< (Q3 + 1.5*IQR))

minimum=min(Top100_no_outliers$popularity)
maximum=max(Top100_no_outliers$popularity)

print(minimum)
print(maximum)

```
```{r}
hist(Top100_no_outliers$popularity)

```
#We will be defining a song being popular as being Top 100 worthy and therefore will establish our cutoff point at 58 based on the histogram above

```{r}
spotify2$is_popular = ifelse(spotify2$popularity >= 58, 1, 0)
```

```{r}
head(spotify2)
```
#We are dropping 'popularity'as its converted into 'is_popular'. We are also dropping 'artist_name' and 'track_name' since we are looking at the anatomy of a popular song and not who sings it or what it's called.
#The goal is to identify songs that will become popular without being affected by the artist's name since we would also like to find songs from up-and-coming artists.

```{r}
spotify3 <- subset(spotify2, select = -c(popularity, artist_name, track_name))
```

```{r}
head(spotify3)
```

```{r}
dim(spotify3)
```
#One Hot Encoding 
```{r}
#read df_ohe_test dataset
spotifydf = read.csv("D:/College/FallSemester/Advanced Analytics/Project/Spotify-Song-Popularity-Prediction-master/data/df_ohe.csv",stringsAsFactors = TRUE)
spotifydf <- spotifydf[ -c(1) ]
spotifydf$is_popular = as.factor(spotifydf$is_popular)
#spotifydf$is_popular = as.factor(spotifydf$is_popular)
dim(spotifydf)
head(spotifydf)

```
#splitting the data into training set and test set
```{r}
library(dplyr)
#remove non numerical column
spotifydf <- subset(spotifydf, select = -c(track_id))
#split dataset
print ("Spotify Training Dataset")
s_train  <- spotifydf %>% dplyr::sample_frac(0.7)
#write.csv(strain, "D:\\College\\FallSemester\\Advanced Analytics\\Project\\Spotify-Song-Popularity-Prediction-master\\data\\strain.csv", row.names=FALSE)
print ("Spotify Testing Dataset")
s_test   <- dplyr::anti_join(spotifydf,s_train)
#write.csv(stest, "D:\\College\\FallSemester\\Advanced Analytics\\Project\\Spotify-Song-Popularity-Prediction-master\\data\\stest.csv", row.names=FALSE)
```


```{r}
dim(s_train)
dim(s_test)
```

#MODELS
#MODEL 1: Logistic Regression
```{r}
# Logistic Regression Model
spotify_log <- glm(is_popular ~., data =s_train, family =binomial)
summary(spotify_log)
```

```{r}
spotify_pred_test <- predict(spotify_log,s_test,type="response")
```
```{r}
#choose the best threshold using ROC
library(ROCR)
ROCR_pred_test <- prediction(spotify_pred_test,s_test$is_popular)
ROCR_perf_test <- performance(ROCR_pred_test,'tpr','fpr')
plot(ROCR_perf_test,colorize=TRUE,print.cutoffs.at=seq(0.1,by=0.1))
```
```{r}
#Confusion Matrix threshold > 0.1
table(s_test$is_popular, as.numeric(spotify_pred_test>0.1)) 
```

```{r}
#Confusion Matrix threshold > 0.2
table(s_test$is_popular, as.numeric(spotify_pred_test>0.2)) 
```

```{r}
varImp(spotify_log,sort=TRUE,n.var=10,main="Top 10 Important Variables")

```


```{r}
#Confusion Matrix threshold > 0.3
table(s_test$is_popular, as.numeric(spotify_pred_test>0.3)) 
```

```{r}
library(randomForest)
srf <- randomForest(is_popular~.,data=s_train)
print(srf)
```


```{r}
#Prediction on testing dataset
library(caret)
predict1<- predict(srf,s_test)
confusionMatrix(predict1,s_test$is_popular)
```
```{r}
#Importance of Variables
varImpPlot(srf,sort=TRUE,n.var=10,main="Top 10 Important Variables")
```

## XGBOOST

### Convert the train and test data into xgboost matrix type.

```{r}
library(xgboost)
library(caret)  

library(magrittr)
library(dplyr)
library(Matrix)
```
```{r}
X_train = data.matrix(s_train[,-53])                  # independent variables for train
y_train = s_train[,53]                                # dependent variables for train
  
X_test = data.matrix(s_test[,-53])                    # independent variables for test
y_test = s_test[,53]                                   # dependent variables for test

# convert the train and test data into xgboost matrix type.
xgboost_train = xgb.DMatrix(data=as.matrix(X_train), label=y_train)
xgboost_test = xgb.DMatrix(data=as.matrix(X_test), label=y_test)
```

```{r}
head(s_train)
```
```{r}
str(xgboost_train)
```
#Parameters
```{r}
nc <- length(unique(y_train))

xgb_params <- list ("objective" = "multi:softprob",
                    "eval_mertic" = "mlogloss",
                    "num_class" = nc)
```

# eXtreme Gradient Boosting Model

```{r}
bst_model <- xgb.train(params = xgb_params,
                       data = xgboost_train,
                       nrounds = 1000,
                       eta = 0.5,
                       max.depth = 3,
                       gamma = 0,
                       subsample = 1,
                       colsample_bytree = 1,
                       missing = NA,
                       seed = 333)
```
# Training & test error plot

```{r}

#bst_model$evaluation_log
#plot(e$iter, e$train_mlogloss, col = 'blue')
#lines(e$iter, e$test_mlogloss, col = 'red')
#bst_model
```
```{r}
# Feature importance
imp <- xgb.importance(colnames(xgboost_train), model = bst_model)
xgb.plot.importance(imp)
```
```{r}
# Prediction & confusion matrix - test data
p <- predict(bst_model, newdata = xgboost_test)
pred <- matrix(p, nrow = nc, ncol = length(p)/nc) %>%
         t() %>%
         data.frame() %>%
         mutate(label = y_test, max_prob = max.col(., "last")-1)
table(Prediction = pred$max_prob, Actual = pred$label)

```
## Accuracy
```{r}
46934/52853

```

```{r}
plot(bst_model)
```

```{r}
#Splitting the file to popular to unpopular dataframes
popular_songs=subset(df_ohe_clean,is_popular=="1")
unpopular_songs=subset(df_ohe_clean,is_popular=="0")


```

```{r}
head(popular_songs)
head(unpopular_songs)
write.csv(popular_songs, "D:\\College\\FallSemester\\Advanced Analytics\\Project\\Spotify-Song-Popularity-Prediction-master\\data\\popular.csv", row.names=FALSE)
write.csv(unpopular_songs, "D:\\College\\FallSemester\\Advanced Analytics\\Project\\Spotify-Song-Popularity-Prediction-master\\data\\unpopular.csv", row.names=FALSE)
```

```{r}
#checking for genre occurence counts for popular songs
s=subset(popular_songs,select = c(11:36))
genrecount<-colSums(s==1)
p=tail(sort(genrecount),5)
barplot(p,main="Top 5 Most Frequent Genres among popular songs",col="blue")


```
```{r}
#checking for genre occurrence counts for unpopular songs
s1=subset(unpopular_songs,select = c(11:36))
genrecount1<-colSums(s1==1)
p1=tail(sort(genrecount1),5)
barplot(p1,main="Top 5 Most Frequent Genres among unpopular songs",col="blue")

```

```{r}
#Acousticness score 
m1<-mean(popular_songs$acousticness)
m2 <- mean(unpopular_songs$acousticness)
m=c(m1,m2)
m
```
```{r}
#Duration score 
m1<-mean(popular_songs$duration_ms)
m2 <- mean(unpopular_songs$duration_ms)
m=c(m1,m2)
m
```
```{r}
#Speechiness score 
m1<-mean(popular_songs$speechiness)
m2 <- mean(unpopular_songs$speechiness)
m=c(m1,m2)
m
```
```{r}
#Valence score 
m1<-mean(popular_songs$valence)
m2 <- mean(unpopular_songs$valence)
m=c(m1,m2)
m
```
```{r}
#IQR of energy of popular track

#find Q1, Q3, and interquartile range for values in column A
Q1 <- quantile(popular_songs$energy, .25)
Q3 <- quantile(popular_songs$energy, .75)
IQR<-IQR(popular_songs$energy)

#only keep rows in dataframe that have values within 1.5*IQR of Q1 and Q3
popular_no_outliers <- subset(popular_songs, popular_songs$energy> (Q1 - 1.5*IQR) & popular_songs$energy< (Q3 + 1.5*IQR))
m1=mean(popular_no_outliers$energy)
#view row and column count of new data frame
dim(popular_songs)
dim(popular_no_outliers) 
```

```{r}
#IQR of energy of unpopular track

#find Q1, Q3, and interquartile range for values in column A
Q1 <- quantile(unpopular_songs$energy, .25)
Q3 <- quantile(unpopular_songs$energy, .75)
IQR<-IQR(unpopular_songs$energy)

#only keep rows in dataframe that have values within 1.5*IQR of Q1 and Q3
unpopular_no_outliers <- subset(unpopular_songs, unpopular_songs$energy> (Q1 - 1.5*IQR) & unpopular_songs$energy< (Q3 + 1.5*IQR))
m2=mean(unpopular_no_outliers$energy)
m2
m=c(m1,m2)
m
#view row and column count of new data frame
dim(unpopular_songs)
dim(unpopular_no_outliers) 
library(RColorBrewer)
coul <- brewer.pal(5, "Set2")
bp=barplot(m,main="Mean Energy Scores for popular and unpopular tracks",col=coul,xlab="Mean Energy Score",ylab="Popularity",horiz=TRUE)
text(bp,pos=3,m)

```

```{r}
#IQR of danceability of popular track

#find Q1, Q3, and interquartile range for values in column A
Q1 <- quantile(popular_songs$danceability, .25)
Q3 <- quantile(popular_songs$danceability, .75)
IQR<-IQR(popular_songs$danceability)

#only keep rows in dataframe that have values within 1.5*IQR of Q1 and Q3
popular_no_outliers <- subset(popular_songs, popular_songs$danceability> (Q1 - 1.5*IQR) & popular_songs$danceability< (Q3 + 1.5*IQR))
m1=mean(popular_no_outliers$danceability)
m1
#view row and column count of new data frame
dim(popular_songs)
dim(popular_no_outliers) 

```



```{r}
#IQR of danceability of unpopular track

#find Q1, Q3, and interquartile range for values in column A
Q1 <- quantile(unpopular_songs$danceability, .25)
Q3 <- quantile(unpopular_songs$danceability, .75)
IQR<-IQR(unpopular_songs$danceability)

#only keep rows in dataframe that have values within 1.5*IQR of Q1 and Q3
unpopular_no_outliers <- subset(unpopular_songs, unpopular_songs$danceability> (Q1 - 1.5*IQR) & unpopular_songs$danceability< (Q3 + 1.5*IQR))
m2=mean(unpopular_no_outliers$danceability)
#view row and column count of new data frame
dim(unpopular_songs)
dim(unpopular_no_outliers) 
m=c(m1,m2)
m
library(RColorBrewer)
coul <- brewer.pal(5, "Set2")
bp=barplot(m,main="Mean Danceability Scores for popular and unpopular tracks",col=coul,xlab="Mean Danceability Score",ylab="Popularity",horiz=TRUE)
text(bp,pos=3,m)

```

```{r}
#IQR of acousticness of popular track

#find Q1, Q3, and interquartile range 
Q1 <- quantile(popular_songs$acousticness, .25)
Q3 <- quantile(popular_songs$acousticness, .75)
IQR<-IQR(popular_songs$acousticness)

#only keep rows in dataframe that have values within 1.5*IQR of Q1 and Q3
popular_no_outliers <- subset(popular_songs, popular_songs$acousticness> (Q1 - 1.5*IQR) & popular_songs$acousticness< (Q3 + 1.5*IQR))
m1=mean(popular_no_outliers$acousticness)
m1
#view row and column count of new data frame
dim(popular_songs)
dim(popular_no_outliers) 

```



```{r}
#IQR of danceability of unpopular track

#find Q1, Q3, and interquartile range for values in column A
Q1 <- quantile(unpopular_songs$acousticness, .25)
Q3 <- quantile(unpopular_songs$acousticness, .75)
IQR<-IQR(unpopular_songs$acousticness)

#only keep rows in dataframe that have values within 1.5*IQR of Q1 and Q3
unpopular_no_outliers <- subset(unpopular_songs, unpopular_songs$acousticness> (Q1 - 1.5*IQR) & unpopular_songs$acousticness< (Q3 + 1.5*IQR))
m2=mean(unpopular_no_outliers$acousticness)
#view row and column count of new data frame
dim(unpopular_songs)
dim(unpopular_no_outliers) 
m=c(m1,m2)
m
library(RColorBrewer)
coul <- brewer.pal(5, "Set2")
bp=barplot(m,main="Mean Acousticness Scores for popular and unpopular tracks",col=coul,xlab="Mean Acousticness Score",ylab="Popularity",horiz=TRUE)
text(bp,pos=3,m)

```

